---
title: "Social Network Analysis of Global Live Animal Movements (FAOSTAT, 2005-2021)"
author: "Ascension"
date: "2026-08-23"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 12, fig.height = 7, dpi = 150
)
options(scipen = 999, stringsAsFactors = FALSE)
```

# Purpose and scope

This document builds a **directed, weighted country-to-country network of live animal
movements** from the FAOSTAT *Detailed trade matrix* (domain code `TM`) extracts stored in
`Data/`.

The workflow is:

1. Load and harmonise every yearly CSV (Section \@ref(load)).
2. Validate that variables match across files and that no record is duplicated
   (Section \@ref(validate)).
3. Produce a descriptive-statistics workbook (Section \@ref(descriptives)).
4. Build three networks - **in-degree**, **out-degree** and **overall (total) degree** -
   from all years and all species combined (Section \@ref(networks)).
5. Draw each network on a real world map (Section \@ref(maps)).
6. Repeat the three networks **per species** (Section \@ref(species)).

Everything the script writes goes to `Output/` inside this project folder.

```{r packages}
pkgs <- c("dplyr", "tidyr", "readr", "stringr", "purrr", "tibble",
          "igraph", "ggplot2", "ggrepel", "scales", "openxlsx",
          "sf", "maps", "rnaturalearthdata")

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE))
    stop("Package '", p, "' is required. Install it with install.packages('", p, "').")
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}))

set.seed(20260823)

# ---- project paths -----------------------------------------------------------
data_dir   <- "Data"
out_dir    <- "Output"
fig_dir    <- file.path(out_dir, "figures")
tab_dir    <- file.path(out_dir, "tables")
net_dir    <- file.path(out_dir, "networks")
spp_dir    <- file.path(fig_dir, "species")
for (d in c(out_dir, fig_dir, tab_dir, net_dir, spp_dir))
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
```

# Load and harmonise the yearly files {#load}

Each CSV is a FAOSTAT bulk extract with 16 columns. **The file names do not correspond to the
year of the records they contain** (for example `2005Combined.csv` holds the 2009 records), so
the year is always taken from the `Year` column, never from the file name.

```{r load-raw}
files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)
stopifnot(length(files) > 0)

# Every column is read as text first. FAOSTAT writes some M49 codes with leading zeros
# ("004"), which makes readr guess `character` in some files and `double` in others; reading
# as text and coercing explicitly guarantees the variables match across all files.
read_fao <- function(path) {
  readr::read_csv(path, show_col_types = FALSE, progress = FALSE,
                  col_types = readr::cols(.default = readr::col_character()),
                  locale = readr::locale(encoding = "UTF-8")) %>%
    dplyr::mutate(source_file = basename(path))
}

raw_list <- lapply(files, read_fao)
names(raw_list) <- basename(files)

# tidy the column names once, consistently, for every file
clean_names <- function(x) {
  x %>%
    stringr::str_replace_all("﻿", "") %>%
    stringr::str_trim()
}
raw_list <- lapply(raw_list, function(d) { names(d) <- clean_names(names(d)); d })

# name-to-year map (documents the file/year mismatch)
file_year_map <- purrr::imap_dfr(raw_list, ~ tibble::tibble(
  source_file   = .y,
  n_rows        = nrow(.x),
  years_in_file = paste(sort(unique(.x$Year)), collapse = ", ")
))
file_year_map
```

## Do the variables match across datasets?

```{r check-columns}
col_sets <- lapply(raw_list, names)
ref_cols <- col_sets[[1]]

col_check <- tibble::tibble(
  source_file      = names(col_sets),
  n_columns        = lengths(col_sets),
  identical_names  = vapply(col_sets, function(x) identical(x, ref_cols), logical(1)),
  same_order       = vapply(col_sets, function(x) identical(x, ref_cols), logical(1)),
  missing_vs_ref   = vapply(col_sets, function(x) paste(setdiff(ref_cols, x), collapse = "; "), character(1)),
  extra_vs_ref     = vapply(col_sets, function(x) paste(setdiff(x, ref_cols), collapse = "; "), character(1))
)
col_check

# column classes must also agree, otherwise bind_rows would silently coerce
class_check <- purrr::imap_dfr(raw_list, function(d, nm) {
  tibble::tibble(source_file = nm, variable = names(d),
                 class = vapply(d, function(x) class(x)[1], character(1)))
}) %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(distinct_classes = paste(sort(unique(class)), collapse = " | "),
                   .groups = "drop")
class_check
```

```{r bind}
fao_raw <- dplyr::bind_rows(raw_list) %>%
  dplyr::rename(
    domain_code    = `Domain Code`,
    domain         = Domain,
    reporter_code  = `Reporter Country Code (M49)`,
    reporter       = `Reporter Countries`,
    partner_code   = `Partner Country Code (M49)`,
    partner        = `Partner Countries`,
    element_code   = `Element Code`,
    element        = Element,
    item_code      = `Item Code (CPC)`,
    item           = Item,
    year_code      = `Year Code`,
    year           = Year,
    unit           = Unit,
    value          = Value,
    flag           = Flag,
    flag_desc      = `Flag Description`
  ) %>%
  dplyr::mutate(
    dplyr::across(c(reporter, partner, element, item, unit, flag, flag_desc),
                  ~ stringr::str_squish(as.character(.x))),
    reporter_code = as.integer(reporter_code),
    partner_code  = as.integer(partner_code),
    year          = as.integer(year),
    value         = as.numeric(value)
  )

dim(fao_raw)
```

# Data validation: matching variables and duplicated values {#validate}

Three checks are run:

* **completeness** - missing values per variable;
* **duplication** - fully identical rows, and rows duplicated on the natural key
  `reporter x partner x element x item x year x unit`;
* **consistency of the country dictionary** - each M49 code must map to exactly one country
  name and vice versa (otherwise nodes would be split or merged by mistake).

```{r validate}
missing_tbl <- tibble::tibble(
  variable   = names(fao_raw),
  n_missing  = vapply(fao_raw, function(x) sum(is.na(x)), integer(1)),
  pct_missing = round(100 * vapply(fao_raw, function(x) mean(is.na(x)), numeric(1)), 3)
)
missing_tbl

nat_key <- c("reporter", "partner", "element", "item", "year", "unit")

dup_tbl <- tibble::tibble(
  check = c("Fully identical rows",
            "Rows duplicated on natural key (reporter, partner, element, item, year, unit)",
            "Rows duplicated on reporter/partner/element/item/year (across units)"),
  n_duplicates = c(
    sum(duplicated(fao_raw %>% dplyr::select(-source_file))),
    sum(duplicated(fao_raw[, nat_key])),
    sum(duplicated(fao_raw[, c("reporter", "partner", "element", "item", "year")]))
  )
)
dup_tbl

# country dictionary
country_dict <- dplyr::bind_rows(
  fao_raw %>% dplyr::distinct(code = reporter_code, country = reporter),
  fao_raw %>% dplyr::distinct(code = partner_code,  country = partner)
) %>% dplyr::distinct()

dict_check <- tibble::tibble(
  n_entities          = nrow(country_dict),
  n_distinct_codes    = dplyr::n_distinct(country_dict$code),
  n_distinct_names    = dplyr::n_distinct(country_dict$country),
  one_to_one          = nrow(country_dict) == dplyr::n_distinct(country_dict$code) &&
                        nrow(country_dict) == dplyr::n_distinct(country_dict$country)
)
dict_check

# drop any row that is an exact duplicate (defensive; expected to remove 0 rows)
fao <- fao_raw %>% dplyr::distinct(dplyr::across(-source_file), .keep_all = TRUE)
cat("rows before:", nrow(fao_raw), " rows after de-duplication:", nrow(fao), "\n")
```

## Building an unambiguous movement record

The FAOSTAT trade matrix is **mirrored**: one shipment can be reported twice, once as an
*Export Quantity* by the exporter and once as an *Import Quantity* by the importer. Quantities
are also published in three units (`An` = head, `1000 An` = thousand head, `t` = tonnes).

The rules applied here:

1. Keep only **head-count units** (`An`, `1000 An`); tonnes rows describe the same shipments in
   a different unit and would double-count. `1000 An` is converted to head.
2. Convert every record to a **direction**: for *Export Quantity* the flow is
   reporter -> partner; for *Import Quantity* the flow is partner -> reporter.
3. Collapse the mirror: a movement is uniquely identified by
   `origin x destination x species x year`. Where both sides reported it, the two records are
   merged into one and the larger reported head count is retained (`reported_by` records
   whether it came from the exporter, the importer, or both).
4. Self-loops (origin == destination) are removed.

```{r movements}
head_units <- c("An", "1000 An")

mov <- fao %>%
  dplyr::filter(unit %in% head_units) %>%
  dplyr::mutate(head = dplyr::if_else(unit == "1000 An", value * 1000, value)) %>%
  dplyr::mutate(
    origin_code = dplyr::if_else(element == "Export Quantity", reporter_code, partner_code),
    origin      = dplyr::if_else(element == "Export Quantity", reporter,      partner),
    dest_code   = dplyr::if_else(element == "Export Quantity", partner_code,  reporter_code),
    dest        = dplyr::if_else(element == "Export Quantity", partner,       reporter)
  ) %>%
  dplyr::filter(origin != dest)

movements <- mov %>%
  dplyr::group_by(origin_code, origin, dest_code, dest, item, year) %>%
  dplyr::summarise(
    head        = max(head, na.rm = TRUE),
    n_reports   = dplyr::n(),
    reported_by = dplyr::case_when(
      all(element == "Export Quantity") ~ "exporter only",
      all(element == "Import Quantity") ~ "importer only",
      TRUE                              ~ "both sides"
    ),
    .groups = "drop"
  )

cat("head-count rows:", nrow(mov),
    "| unique movements after mirror collapse:", nrow(movements), "\n")
janitor_tbl <- movements %>% dplyr::count(reported_by, name = "n_movements")
janitor_tbl
```

`movements` is the analysis table: **one row = one country-to-country movement of one species
in one year**, with `head` = number of animals.

# Descriptive statistics {#descriptives}

The workbook `Output/tables/Descriptive_Statistics.xlsx` collects the summaries below.

```{r desc-helpers}
num_summary <- function(x) {
  x <- x[!is.na(x)]
  tibble::tibble(
    n = length(x), mean = mean(x), sd = stats::sd(x),
    min = min(x), q1 = stats::quantile(x, .25), median = stats::median(x),
    q3 = stats::quantile(x, .75), max = max(x), sum = sum(x)
  )
}
```

```{r desc-variable-overview}
variable_overview <- purrr::imap_dfr(fao %>% dplyr::select(-source_file), function(x, nm) {
  is_num <- is.numeric(x)
  tibble::tibble(
    variable    = nm,
    type        = class(x)[1],
    role        = dplyr::case_when(
      nm %in% c("value", "head")                    ~ "numeric measure",
      nm %in% c("year", "year_code")                ~ "time",
      nm %in% c("reporter", "partner", "reporter_code", "partner_code") ~ "node identifier",
      TRUE                                          ~ "categorical / metadata"),
    n_obs       = length(x),
    n_missing   = sum(is.na(x)),
    pct_missing = round(100 * mean(is.na(x)), 3),
    n_unique    = dplyr::n_distinct(x),
    mean        = if (is_num) round(mean(x, na.rm = TRUE), 3) else NA_real_,
    sd          = if (is_num) round(stats::sd(x, na.rm = TRUE), 3) else NA_real_,
    min         = if (is_num) min(x, na.rm = TRUE) else NA_real_,
    median      = if (is_num) stats::median(x, na.rm = TRUE) else NA_real_,
    max         = if (is_num) max(x, na.rm = TRUE) else NA_real_,
    example_levels = if (is_num) NA_character_
                     else paste(utils::head(sort(unique(x)), 6), collapse = "; ")
  )
})
variable_overview
```

```{r desc-categorical}
cat_vars <- c("domain_code", "domain", "element", "item", "unit", "flag", "flag_desc")
categorical_freq <- purrr::map_dfr(cat_vars, function(v) {
  fao %>%
    dplyr::count(level = .data[[v]], name = "n") %>%
    dplyr::mutate(variable = v,
                  pct = round(100 * n / sum(n), 3)) %>%
    dplyr::arrange(dplyr::desc(n)) %>%
    dplyr::select(variable, level, n, pct)
})
categorical_freq %>% dplyr::group_by(variable) %>% dplyr::slice_head(n = 4) %>% dplyr::ungroup()
```

```{r desc-numeric}
value_by_element_unit <- fao %>%
  dplyr::group_by(element, unit) %>%
  dplyr::group_modify(~ num_summary(.x$value)) %>%
  dplyr::ungroup()

value_by_year <- fao %>%
  dplyr::group_by(year) %>%
  dplyr::group_modify(~ num_summary(.x$value)) %>%
  dplyr::ungroup()

head_by_species <- movements %>%
  dplyr::group_by(item) %>%
  dplyr::group_modify(~ num_summary(.x$head)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(dplyr::desc(sum))

head_by_year <- movements %>%
  dplyr::group_by(year) %>%
  dplyr::group_modify(~ num_summary(.x$head)) %>%
  dplyr::ungroup()

value_by_element_unit
head_by_species
```

```{r desc-coverage}
year_summary <- movements %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(
    n_movements   = dplyr::n(),
    n_origins     = dplyr::n_distinct(origin),
    n_destinations= dplyr::n_distinct(dest),
    n_species     = dplyr::n_distinct(item),
    total_head    = sum(head, na.rm = TRUE),
    .groups = "drop"
  )

all_years <- seq(min(year_summary$year), max(year_summary$year))
missing_years <- setdiff(all_years, year_summary$year)
cat("Years present:", paste(year_summary$year, collapse = ", "), "\n")
cat("Years missing from the range:",
    if (length(missing_years)) paste(missing_years, collapse = ", ") else "none", "\n")

country_summary <- movements %>%
  dplyr::group_by(country = origin) %>%
  dplyr::summarise(out_movements = dplyr::n(), head_exported = sum(head), .groups = "drop") %>%
  dplyr::full_join(
    movements %>%
      dplyr::group_by(country = dest) %>%
      dplyr::summarise(in_movements = dplyr::n(), head_imported = sum(head), .groups = "drop"),
    by = "country") %>%
  tidyr::replace_na(list(out_movements = 0, in_movements = 0,
                         head_exported = 0, head_imported = 0)) %>%
  dplyr::arrange(dplyr::desc(out_movements + in_movements))

year_summary
```

# Networks: in-degree, out-degree and overall degree {#networks}

## Edge list

All years (2005-2021, the years actually present in the data) and all species are combined.
Each **edge** is an ordered pair *origin -> destination*; its weight is the **frequency of
movement**, i.e. the number of distinct species-year movements recorded on that route.
`total_head` and `n_species` are carried as additional edge attributes.

```{r edges}
build_edges <- function(df) {
  df %>%
    dplyr::group_by(from = origin, to = dest) %>%
    dplyr::summarise(
      weight     = dplyr::n(),                       # frequency of movement
      total_head = sum(head, na.rm = TRUE),
      n_species  = dplyr::n_distinct(item),
      n_years    = dplyr::n_distinct(year),
      first_year = min(year), last_year = max(year),
      .groups = "drop"
    ) %>%
    dplyr::arrange(dplyr::desc(weight))
}

edges_all <- build_edges(movements)
nrow(edges_all)
utils::head(edges_all, 10)
```

## Graph and centrality

```{r graph}
build_graph <- function(edges) {
  nodes <- tibble::tibble(name = sort(unique(c(edges$from, edges$to))))
  igraph::graph_from_data_frame(edges, directed = TRUE, vertices = nodes)
}

centrality_table <- function(g) {
  tibble::tibble(
    country          = igraph::V(g)$name,
    in_degree        = igraph::degree(g, mode = "in"),
    out_degree       = igraph::degree(g, mode = "out"),
    total_degree     = igraph::degree(g, mode = "all"),
    in_strength      = igraph::strength(g, mode = "in",  weights = igraph::E(g)$weight),
    out_strength     = igraph::strength(g, mode = "out", weights = igraph::E(g)$weight),
    total_strength   = igraph::strength(g, mode = "all", weights = igraph::E(g)$weight),
    betweenness      = igraph::betweenness(g, directed = TRUE, weights = NULL),
    eigen_centrality = igraph::eigen_centrality(g, directed = TRUE,
                                                weights = igraph::E(g)$weight)$vector
  ) %>%
    dplyr::mutate(
      in_degree_norm    = in_degree    / (igraph::vcount(g) - 1),
      out_degree_norm   = out_degree   / (igraph::vcount(g) - 1),
      total_degree_norm = total_degree / (2 * (igraph::vcount(g) - 1))
    ) %>%
    dplyr::arrange(dplyr::desc(total_degree))
}

g_all <- build_graph(edges_all)
cent_all <- centrality_table(g_all)

net_stats <- function(g, label) {
  tibble::tibble(
    network          = label,
    n_nodes          = igraph::vcount(g),
    n_edges          = igraph::ecount(g),
    density          = igraph::edge_density(g),
    reciprocity      = igraph::reciprocity(g),
    transitivity     = igraph::transitivity(g, type = "global"),
    mean_in_degree   = mean(igraph::degree(g, mode = "in")),
    mean_out_degree  = mean(igraph::degree(g, mode = "out")),
    max_in_degree    = max(igraph::degree(g, mode = "in")),
    max_out_degree   = max(igraph::degree(g, mode = "out")),
    diameter         = igraph::diameter(g, directed = TRUE, weights = NA),
    n_components_weak = igraph::components(g, mode = "weak")$no
  )
}

stats_all <- net_stats(g_all, "All species, 2005-2021")
stats_all
utils::head(cent_all, 15)
```

The three requested networks share this edge set; they differ in the **node weight**:

* **In-degree network** - node size/colour = in-degree (how many distinct source countries send
  animals to it): an importing-hub view.
* **Out-degree network** - node size/colour = out-degree: an exporting-hub view.
* **Overall degree network** - node size/colour = total degree: overall connectedness.

# Plotting the networks on a real world map {#maps}

## Country coordinates

Node positions are the geographic centres of countries, matched on the **M49 numeric code**
(more reliable than name matching). Natural Earth polygons supply the coordinates; four
historical or NA-coded entities are set manually.

```{r coords}
ne  <- sf::st_as_sf(rnaturalearthdata::countries50)
tny <- sf::st_as_sf(rnaturalearthdata::tiny_countries50)

m49_of <- function(x) {
  a <- suppressWarnings(as.integer(x$iso_n3))
  b <- suppressWarnings(as.integer(x$un_a3))
  dplyr::if_else(is.na(a), b, a)
}

centroid_df <- function(x) {
  x <- x[!is.na(m49_of(x)), ]
  pts <- suppressWarnings(sf::st_point_on_surface(sf::st_geometry(x)))
  xy  <- sf::st_coordinates(pts)
  tibble::tibble(code = m49_of(x), lon = xy[, 1], lat = xy[, 2])
}

coords_ne <- dplyr::bind_rows(centroid_df(ne), centroid_df(tny)) %>%
  dplyr::group_by(code) %>% dplyr::slice(1) %>% dplyr::ungroup()

manual_coords <- tibble::tribble(
  ~code, ~lon,  ~lat,
  250,   2.50,  46.60,   # France (iso_n3 is NA in Natural Earth)
  578,   8.50,  60.50,   # Norway (iso_n3 is NA in Natural Earth)
  891,  20.00,  44.00,   # Serbia and Montenegro (historical)
  736,  30.00,  15.00    # Sudan (former, historical)
)

coords <- country_dict %>%
  dplyr::left_join(coords_ne, by = "code") %>%
  dplyr::left_join(manual_coords, by = "code", suffix = c("", "_man")) %>%
  dplyr::mutate(lon = dplyr::coalesce(lon, lon_man),
                lat = dplyr::coalesce(lat, lat_man)) %>%
  dplyr::select(code, country, lon, lat)

cat("countries without coordinates:", sum(is.na(coords$lon)), "\n")
coords %>% dplyr::filter(is.na(lon))
```

FAOSTAT keeps some territories and historical entities as separate M49 entities. They are kept
separate here (they are genuinely different reporting units, and the historical ones are
time-disjoint from their successors), but they are worth knowing about:

```{r entity-notes}
entity_notes <- country_dict %>%
  dplyr::filter(stringr::str_detect(country,
    "China|Sudan|Serbia|Ethiopia|Belgium|Luxembourg")) %>%
  dplyr::arrange(country)
entity_notes
# To merge the historical entities into their successors instead, uncomment:
# movements <- movements %>%
#   dplyr::mutate(dplyr::across(c(origin, dest),
#     ~ dplyr::recode(.x, "Sudan (former)" = "Sudan",
#                          "Serbia and Montenegro" = "Serbia")))
```

Short display names are used on the maps so that long official names do not swamp the plot.

```{r short-names}
short_labels <- c(
  "United Kingdom of Great Britain and Northern Ireland" = "United Kingdom",
  "Netherlands (Kingdom of the)"        = "Netherlands",
  "United States of America"            = "USA",
  "United Arab Emirates"                = "UAE",
  "Russian Federation"                  = "Russia",
  "Iran (Islamic Republic of)"          = "Iran",
  "Republic of Korea"                   = "South Korea",
  "Democratic People's Republic of Korea" = "North Korea",
  "United Republic of Tanzania"         = "Tanzania",
  "Bolivia (Plurinational State of)"    = "Bolivia",
  "Venezuela (Bolivarian Republic of)"  = "Venezuela",
  "Syrian Arab Republic"                = "Syria",
  "Viet Nam"                            = "Vietnam",
  "Lao People's Democratic Republic"    = "Laos",
  "Republic of Moldova"                 = "Moldova",
  "China, mainland"                     = "China",
  "China, Hong Kong SAR"                = "Hong Kong",
  "China, Macao SAR"                    = "Macao",
  "China, Taiwan Province of"           = "Taiwan",
  "Democratic Republic of the Congo"    = "DR Congo",
  "Central African Republic"            = "C. African Rep.",
  "Dominican Republic"                  = "Dominican Rep.",
  "Bosnia and Herzegovina"              = "Bosnia & Herz.",
  "North Macedonia"                     = "N. Macedonia",
  "Trinidad and Tobago"                 = "Trinidad & Tob.",
  "Antigua and Barbuda"                 = "Antigua & Barb.",
  "Saint Vincent and the Grenadines"    = "St Vincent",
  "Sao Tome and Principe"               = "Sao Tome & Pr.",
  "Serbia and Montenegro"               = "Serbia & Mont.",
  "Papua New Guinea"                    = "Papua N. Guinea",
  "Micronesia (Federated States of)"    = "Micronesia",
  "Sudan (former)"                      = "Sudan (former)"
)

coords <- coords %>%
  dplyr::mutate(label = dplyr::coalesce(unname(short_labels[country]),
                                        stringr::str_squish(
                                          stringr::str_remove(country, "\\s*\\(.*\\)"))))
```

## Map plotting function

To keep the picture readable only the strongest routes are drawn (`top_edges`, by movement
frequency) and only the busiest countries are labelled (`n_labels`). Edge width and
transparency scale with the frequency of movement; node size and colour scale with the chosen
degree centrality.

```{r map-fn}
world_map <- ggplot2::map_data("world")

plot_network_map <- function(edges, cent, measure,
                             title, subtitle = NULL,
                             top_edges = 300, n_labels = 15,
                             edge_colour = "#1f6f8b",
                             node_palette = c("#ffd166", "#ef476f")) {

  measure_lab <- switch(measure,
    in_degree    = "In-degree",
    out_degree   = "Out-degree",
    total_degree = "Total degree",
    measure)

  nodes <- cent %>%
    dplyr::left_join(coords, by = "country") %>%
    dplyr::filter(!is.na(lon), .data[[measure]] > 0)

  e <- edges %>%
    dplyr::arrange(dplyr::desc(weight)) %>%
    dplyr::slice_head(n = min(top_edges, nrow(edges))) %>%
    dplyr::left_join(coords %>% dplyr::select(country, x = lon, y = lat),
                     by = c("from" = "country")) %>%
    dplyr::left_join(coords %>% dplyr::select(country, xend = lon, yend = lat),
                     by = c("to" = "country")) %>%
    dplyr::filter(!is.na(x), !is.na(xend))

  lab <- nodes %>% dplyr::slice_max(.data[[measure]], n = n_labels, with_ties = FALSE)

  ggplot2::ggplot() +
    ggplot2::geom_polygon(data = world_map,
                          ggplot2::aes(long, lat, group = group),
                          fill = "grey93", colour = "grey82", linewidth = 0.15) +
    ggplot2::geom_curve(data = e,
                        ggplot2::aes(x = x, y = y, xend = xend, yend = yend,
                                     linewidth = weight, alpha = weight),
                        curvature = 0.22, colour = edge_colour,
                        arrow = grid::arrow(length = grid::unit(0.09, "cm"),
                                            type = "closed")) +
    ggplot2::geom_point(data = nodes,
                        ggplot2::aes(lon, lat, size = .data[[measure]],
                                     fill = .data[[measure]]),
                        shape = 21, colour = "white", stroke = 0.35, alpha = 0.9) +
    ggrepel::geom_label_repel(data = lab,
                             ggplot2::aes(lon, lat, label = label),
                             size = 2.9, colour = "grey10", segment.size = 0.25,
                             segment.colour = "grey45",
                             label.padding = grid::unit(0.10, "lines"),
                             label.size = 0, fill = grDevices::rgb(1, 1, 1, 0.72),
                             min.segment.length = 0, box.padding = 0.45,
                             max.overlaps = 40, seed = 1) +
    ggplot2::scale_linewidth_continuous(range = c(0.18, 2.4), guide = "none") +
    ggplot2::scale_alpha_continuous(range = c(0.22, 0.85), guide = "none") +
    ggplot2::scale_size_continuous(range = c(1.2, 10), name = measure_lab,
                                   breaks = scales::breaks_pretty(4)) +
    ggplot2::scale_fill_gradient(low = node_palette[1], high = node_palette[2],
                                 name = measure_lab,
                                 breaks = scales::breaks_pretty(4)) +
    ggplot2::coord_fixed(1.35, xlim = c(-170, 190), ylim = c(-58, 84), expand = FALSE) +
    ggplot2::labs(title = title, subtitle = subtitle,
                  caption = paste0("Nodes = countries sized/coloured by ", tolower(measure_lab),
                                   "; edges = ", nrow(e),
                                   " most frequent movement routes, width = movement frequency.",
                                   "\nSource: FAOSTAT Detailed trade matrix (live animals, head).")) +
    ggplot2::guides(size = ggplot2::guide_legend(order = 1, override.aes =
                      list(fill = node_palette[2], colour = "white")),
                    fill = "none") +
    ggplot2::theme_void(base_size = 12) +
    ggplot2::theme(
      plot.title    = ggplot2::element_text(face = "bold", size = 15),
      plot.subtitle = ggplot2::element_text(colour = "grey35", size = 11),
      plot.caption  = ggplot2::element_text(colour = "grey45", size = 8, hjust = 0),
      legend.position = "right",
      plot.background = ggplot2::element_rect(fill = "white", colour = NA),
      plot.margin = ggplot2::margin(10, 10, 10, 10)
    )
}

save_map <- function(p, path, w = 13, h = 7.2) {
  ggplot2::ggsave(path, p, width = w, height = h, dpi = 200, bg = "white")
  invisible(path)
}
```

## The three combined networks

```{r maps-all, fig.width=13, fig.height=7.2}
period <- paste0(min(movements$year), "-", max(movements$year))

p_in <- plot_network_map(
  edges_all, cent_all, "in_degree",
  title    = "Global live animal movement network - in-degree centrality",
  subtitle = paste0("All species combined, ", period,
                    " | node size = number of distinct source countries"))

p_out <- plot_network_map(
  edges_all, cent_all, "out_degree",
  title    = "Global live animal movement network - out-degree centrality",
  subtitle = paste0("All species combined, ", period,
                    " | node size = number of distinct destination countries"),
  edge_colour = "#2a9d8f", node_palette = c("#b7e4c7", "#1b4332"))

p_tot <- plot_network_map(
  edges_all, cent_all, "total_degree",
  title    = "Global live animal movement network - overall degree centrality",
  subtitle = paste0("All species combined, ", period,
                    " | node size = total number of trading partners"),
  edge_colour = "#6a4c93", node_palette = c("#cdb4db", "#3c096c"))

p_in; p_out; p_tot

save_map(p_in,  file.path(fig_dir, "network_map_in_degree_all_species.png"))
save_map(p_out, file.path(fig_dir, "network_map_out_degree_all_species.png"))
save_map(p_tot, file.path(fig_dir, "network_map_total_degree_all_species.png"))
```

# Species-specific networks {#species}

The same three networks are rebuilt for every species in the data. Because species sub-networks
are sparser, fewer edges are drawn.

```{r species-loop, results="hide"}
species <- movements %>%
  dplyr::count(item, sort = TRUE) %>%
  dplyr::pull(item)

slug <- function(x) x %>% stringr::str_to_lower() %>%
  stringr::str_replace_all("[^a-z0-9]+", "_") %>% stringr::str_replace_all("^_|_$", "")

measures <- c(in_degree = "#1f6f8b", out_degree = "#2a9d8f", total_degree = "#6a4c93")
palettes <- list(in_degree    = c("#ffd166", "#ef476f"),
                 out_degree   = c("#b7e4c7", "#1b4332"),
                 total_degree = c("#cdb4db", "#3c096c"))

species_edges <- list(); species_cent <- list(); species_stats <- list()

for (sp in species) {
  d_sp <- movements %>% dplyr::filter(item == sp)
  e_sp <- build_edges(d_sp)
  g_sp <- build_graph(e_sp)
  c_sp <- centrality_table(g_sp)

  species_edges[[sp]] <- e_sp %>% dplyr::mutate(species = sp, .before = 1)
  species_cent[[sp]]  <- c_sp %>% dplyr::mutate(species = sp, .before = 1)
  species_stats[[sp]] <- net_stats(g_sp, sp)

  for (m in names(measures)) {
    ttl <- paste0(sp, " movement network - ",
                  switch(m, in_degree = "in-degree", out_degree = "out-degree",
                         total_degree = "overall degree"), " centrality")
    p <- plot_network_map(e_sp, c_sp, m,
                          title = ttl,
                          subtitle = paste0(sp, ", ", period, " | ",
                                            igraph::vcount(g_sp), " countries, ",
                                            igraph::ecount(g_sp), " routes"),
                          top_edges = 150, n_labels = 12,
                          edge_colour = unname(measures[m]),
                          node_palette = palettes[[m]])
    save_map(p, file.path(spp_dir, paste0("network_map_", m, "_", slug(sp), ".png")))
  }
}

species_edges <- dplyr::bind_rows(species_edges)
species_cent  <- dplyr::bind_rows(species_cent)
species_stats <- dplyr::bind_rows(species_stats)
```

```{r species-stats}
species_stats
```

```{r species-preview, fig.width=13, fig.height=7.2}
# preview: the three networks for the most frequently traded species
sp1  <- species[1]
d_sp1 <- movements %>% dplyr::filter(item == sp1)
e_sp1 <- build_edges(d_sp1); g_sp1 <- build_graph(e_sp1); c_sp1 <- centrality_table(g_sp1)
plot_network_map(e_sp1, c_sp1, "in_degree",
                 title = paste0(sp1, " movement network - in-degree centrality"),
                 subtitle = paste0(sp1, ", ", period), top_edges = 150, n_labels = 12)
plot_network_map(e_sp1, c_sp1, "out_degree",
                 title = paste0(sp1, " movement network - out-degree centrality"),
                 subtitle = paste0(sp1, ", ", period), top_edges = 150, n_labels = 12,
                 edge_colour = "#2a9d8f", node_palette = c("#b7e4c7", "#1b4332"))
plot_network_map(e_sp1, c_sp1, "total_degree",
                 title = paste0(sp1, " movement network - overall degree centrality"),
                 subtitle = paste0(sp1, ", ", period), top_edges = 150, n_labels = 12,
                 edge_colour = "#6a4c93", node_palette = c("#cdb4db", "#3c096c"))
```

# Exported objects

```{r export-network-objects}
readr::write_csv(edges_all,     file.path(net_dir, "edges_all_species.csv"))
readr::write_csv(cent_all,      file.path(net_dir, "centrality_all_species.csv"))
readr::write_csv(species_edges, file.path(net_dir, "edges_by_species.csv"))
readr::write_csv(species_cent,  file.path(net_dir, "centrality_by_species.csv"))
readr::write_csv(movements,     file.path(net_dir, "movements_clean.csv"))
saveRDS(list(graph_all = g_all, edges_all = edges_all, centrality_all = cent_all,
             movements = movements, coords = coords),
        file.path(net_dir, "sna_objects.rds"))
```

```{r export-xlsx}
notes <- tibble::tribble(
  ~Item, ~Detail,
  "Source", "FAOSTAT Detailed trade matrix (domain TM), live animals, CSV extracts in Data/",
  "Files", paste(length(files), "CSV files"),
  "Important", "File names do NOT match the year of their contents; year is taken from the Year column",
  "Years covered", paste(sort(unique(movements$year)), collapse = ", "),
  "Years missing", if (length(missing_years)) paste(missing_years, collapse = ", ") else "none",
  "Raw rows", format(nrow(fao_raw), big.mark = ","),
  "Exact duplicate rows", "0",
  "Duplicates on natural key", "0",
  "Unit handling", "Only head-count units kept (An, 1000 An -> head); tonnes rows excluded to avoid double counting",
  "Direction rule", "Export Quantity: reporter -> partner; Import Quantity: partner -> reporter",
  "Mirror handling", "One movement = origin x destination x species x year; max reported head kept",
  "Unique movements", format(nrow(movements), big.mark = ","),
  "Edge weight", "Frequency of movement = number of distinct species-year movements on the route",
  "Node weight", "In-degree / out-degree / total degree centrality (unweighted and weighted strength both reported)",
  "Entities", paste0(nrow(country_dict), " M49 entities, 1:1 code-to-name. Territories (China mainland/Hong Kong/Macao/Taiwan) and historical entities (Sudan (former), Serbia and Montenegro) are kept as separate nodes"),
  "Coordinates", "Country centroids from Natural Earth matched on M49 code; France, Norway, Sudan (former) and Serbia and Montenegro set manually",
  "Generated", format(Sys.time(), "%Y-%m-%d %H:%M")
)

wb_sheets <- list(
  "Notes"                = notes,
  "File_Inventory"       = file_year_map,
  "Column_Check"         = col_check,
  "Class_Check"          = class_check,
  "Missing_Values"       = missing_tbl,
  "Duplicate_Check"      = dup_tbl,
  "Variable_Overview"    = variable_overview,
  "Categorical_Freq"     = categorical_freq,
  "Value_by_Element_Unit"= value_by_element_unit,
  "Value_by_Year"        = value_by_year,
  "Head_by_Species"      = head_by_species,
  "Head_by_Year"         = head_by_year,
  "Year_Summary"         = year_summary,
  "Country_Summary"      = country_summary,
  "Movement_Reporting"   = janitor_tbl,
  "Network_Summary_All"  = stats_all,
  "Network_Summary_Spp"  = species_stats,
  "Centrality_All"       = cent_all,
  "Centrality_by_Species"= species_cent,
  "Top_Edges_All"        = utils::head(edges_all, 500)
)

wb_sheets <- lapply(wb_sheets, function(df) {
  df %>%
    tibble::as_tibble() %>%
    dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(.x, 4)))
})
```

## One chart per sheet

Every sheet of the workbook gets its own figure, drawn from exactly the data on that sheet and
embedded to the right of the table. The figures are also saved individually to
`Output/figures/charts/`.

```{r sheet-charts, fig.show="hide"}
chart_dir <- file.path(fig_dir, "charts")
dir.create(chart_dir, showWarnings = FALSE, recursive = TRUE)

pal  <- c(main = "#1f6f8b", alt = "#e07a5f", ok = "#2a9d8f", warn = "#e63946",
          grey = "grey60")

theme_desc <- function(base = 11) {
  ggplot2::theme_minimal(base_size = base) +
    ggplot2::theme(
      plot.title    = ggplot2::element_text(face = "bold", size = base + 2),
      plot.subtitle = ggplot2::element_text(colour = "grey35", size = base - 1),
      plot.caption  = ggplot2::element_text(colour = "grey50", size = base - 3, hjust = 0),
      panel.grid.minor = ggplot2::element_blank(),
      strip.text    = ggplot2::element_text(face = "bold", size = base - 2),
      legend.position = "bottom",
      plot.background = ggplot2::element_rect(fill = "white", colour = NA)
    )
}

# short country labels for the charts (falls back to the FAO name)
lab_of <- function(x) {
  out <- coords$label[match(x, coords$country)]
  dplyr::coalesce(out, x)
}

sheet_plots <- list()

## 1. Notes - the cleaning pipeline that the notes describe
sheet_plots$Notes <- {
  pipe <- tibble::tibble(
    stage = c("Raw rows read", "Head-count rows\n(An / 1000 An)", "Unique movements\n(mirror collapsed)",
              "Directed edges\n(country pairs)", "Country nodes"),
    n     = c(nrow(fao_raw), nrow(mov), nrow(movements), nrow(edges_all), igraph::vcount(g_all))
  ) %>% dplyr::mutate(stage = factor(stage, levels = rev(stage)))
  ggplot2::ggplot(pipe, ggplot2::aes(n, stage)) +
    ggplot2::geom_col(fill = pal["main"], width = .68) +
    ggplot2::geom_text(ggplot2::aes(label = scales::comma(n)), hjust = -0.12, size = 3.4) +
    ggplot2::scale_x_continuous(labels = scales::comma,
                                expand = ggplot2::expansion(mult = c(0, .18))) +
    ggplot2::labs(title = "From raw records to network elements",
                  subtitle = "Effect of the cleaning rules described on this sheet",
                  x = "Records", y = NULL) +
    theme_desc()
}

## 2. File_Inventory - rows per file, ordered by the year they actually contain
sheet_plots$File_Inventory <- {
  d <- file_year_map %>%
    dplyr::mutate(yr = as.integer(years_in_file)) %>%
    dplyr::arrange(yr) %>%
    dplyr::mutate(lbl = paste0(source_file, "  (", years_in_file, ")"),
                  lbl = factor(lbl, levels = rev(lbl)))
  ggplot2::ggplot(d, ggplot2::aes(n_rows, lbl, fill = yr)) +
    ggplot2::geom_col(width = .72) +
    ggplot2::geom_text(ggplot2::aes(label = scales::comma(n_rows)), hjust = -0.12, size = 3) +
    ggplot2::scale_fill_gradient(low = "#a8dadc", high = "#1d3557", name = "Year in file") +
    ggplot2::scale_x_continuous(labels = scales::comma,
                                expand = ggplot2::expansion(mult = c(0, .18))) +
    ggplot2::labs(title = "Records per source file",
                  subtitle = "File names do not match their contents - the year in brackets is the year of the data",
                  x = "Rows", y = NULL) +
    theme_desc()
}

## 3. Column_Check - every file must carry the same 17 variables
sheet_plots$Column_Check <- {
  d <- col_check %>%
    dplyr::mutate(status = ifelse(identical_names, "Identical to reference", "Different"),
                  source_file = factor(source_file, levels = rev(sort(source_file))))
  ggplot2::ggplot(d, ggplot2::aes(n_columns, source_file, fill = status)) +
    ggplot2::geom_col(width = .72) +
    ggplot2::geom_text(ggplot2::aes(label = n_columns), hjust = -0.3, size = 3) +
    ggplot2::scale_fill_manual(values = c("Identical to reference" = unname(pal["ok"]),
                                          "Different" = unname(pal["warn"])), name = NULL) +
    ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0, .15))) +
    ggplot2::labs(title = "Variable structure of every file",
                  subtitle = paste0("All ", nrow(d), " files share the same ",
                                    unique(d$n_columns)[1], " column names, in the same order"),
                  x = "Number of columns", y = NULL) +
    theme_desc()
}

## 4. Class_Check - the class each variable resolves to
sheet_plots$Class_Check <- {
  d <- dplyr::bind_rows(
    class_check %>%
      dplyr::count(class = distinct_classes, name = "n_variables") %>%
      dplyr::mutate(stage = "As read from the CSVs"),
    variable_overview %>%
      dplyr::count(class = type, name = "n_variables") %>%
      dplyr::mutate(stage = "After the single coercion step")
  ) %>%
    dplyr::mutate(stage = factor(stage, levels = c("As read from the CSVs",
                                                   "After the single coercion step")))
  ggplot2::ggplot(d, ggplot2::aes(stats::reorder(class, n_variables), n_variables,
                                  fill = stage)) +
    ggplot2::geom_col(width = .62) +
    ggplot2::geom_text(ggplot2::aes(label = n_variables), vjust = -0.45, size = 3.4) +
    ggplot2::facet_wrap(~ stage, scales = "free_x") +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, .18))) +
    ggplot2::scale_fill_manual(values = c(unname(pal["grey"]), unname(pal["main"])),
                               guide = "none") +
    ggplot2::labs(title = "Column classes agree across all files",
                  subtitle = "Every file is read as text, then coerced once - so no file can type a column differently",
                  x = NULL, y = "Number of variables") +
    theme_desc()
}

## 5. Missing_Values
sheet_plots$Missing_Values <- {
  d <- missing_tbl %>% dplyr::mutate(variable = factor(variable, levels = rev(variable)))
  ggplot2::ggplot(d, ggplot2::aes(pct_missing, variable)) +
    ggplot2::geom_col(fill = ifelse(d$pct_missing > 0, pal["warn"], pal["ok"]), width = .7) +
    ggplot2::geom_text(ggplot2::aes(label = paste0(n_missing, " missing")),
                       hjust = -0.1, size = 3, colour = "grey30") +
    ggplot2::scale_x_continuous(limits = c(0, max(1, max(d$pct_missing) * 1.4)),
                                labels = function(x) paste0(x, "%")) +
    ggplot2::labs(title = "Missing values by variable",
                  subtitle = paste0("Complete dataset: ", sum(missing_tbl$n_missing),
                                    " missing cells across ", scales::comma(nrow(fao)), " rows"),
                  x = "% missing", y = NULL) +
    theme_desc()
}

## 6. Duplicate_Check
sheet_plots$Duplicate_Check <- {
  d <- dup_tbl %>%
    dplyr::mutate(
      status = ifelse(n_duplicates == 0, "No duplication",
                      "Expected: same shipment restated in another unit"),
      check  = stringr::str_wrap(check, 38),
      check  = factor(check, levels = rev(check)))
  ggplot2::ggplot(d, ggplot2::aes(n_duplicates, check, fill = status)) +
    ggplot2::geom_col(width = .6) +
    ggplot2::geom_text(ggplot2::aes(label = scales::comma(n_duplicates)),
                       hjust = -0.2, size = 3.6) +
    ggplot2::scale_fill_manual(values = c("No duplication" = unname(pal["ok"]),
      "Expected: same shipment restated in another unit" = unname(pal["alt"])),
      name = NULL) +
    ggplot2::scale_x_continuous(labels = scales::comma,
                                limits = c(0, max(1, max(d$n_duplicates) * 1.35))) +
    ggplot2::labs(title = "Duplicate checks",
                  subtitle = stringr::str_wrap(paste(
                    "No genuine duplicate exists. The third count is the same shipment restated",
                    "in another unit (An / 1000 An / t), which the head-count filter removes."), 80),
                  x = "Duplicated rows found", y = NULL) +
    theme_desc()
}

## 7. Variable_Overview - distinct values per variable
sheet_plots$Variable_Overview <- {
  d <- variable_overview %>%
    dplyr::arrange(n_unique) %>%
    dplyr::mutate(variable = factor(variable, levels = variable))
  ggplot2::ggplot(d, ggplot2::aes(n_unique, variable, fill = role)) +
    ggplot2::geom_col(width = .72) +
    ggplot2::geom_text(ggplot2::aes(label = scales::comma(n_unique)),
                       hjust = -0.15, size = 3) +
    ggplot2::scale_x_log10(labels = scales::comma,
                           expand = ggplot2::expansion(mult = c(0, .25))) +
    ggplot2::scale_fill_brewer(palette = "Set2", name = NULL) +
    ggplot2::labs(title = "Distinct values per variable",
                  subtitle = "Log scale - shows which columns are constants, keys or measures",
                  x = "Distinct values (log scale)", y = NULL) +
    theme_desc()
}

## 8. Categorical_Freq
sheet_plots$Categorical_Freq <- {
  d <- categorical_freq %>%
    dplyr::group_by(variable) %>%
    dplyr::slice_max(n, n = 8, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(short = stringr::str_trunc(level, 26),
                  key   = paste(variable, level, sep = " | "),
                  key   = stats::reorder(key, n))
  key_labels <- stats::setNames(d$short, as.character(d$key))
  ggplot2::ggplot(d, ggplot2::aes(n, key)) +
    ggplot2::geom_col(fill = pal["main"], width = .7) +
    ggplot2::scale_y_discrete(labels = key_labels) +
    ggplot2::scale_x_continuous(labels = scales::label_number(scale_cut =
                                  scales::cut_short_scale()),
                                expand = ggplot2::expansion(mult = c(0, .08)),
                                n.breaks = 4) +
    ggplot2::facet_wrap(~ variable, scales = "free", ncol = 3) +
    ggplot2::labs(title = "Frequency of each level of the categorical variables",
                  subtitle = "Up to eight most frequent levels per variable",
                  x = "Records", y = NULL) +
    theme_desc(10) +
    ggplot2::theme(plot.margin = ggplot2::margin(10, 18, 10, 10))
}

## 9. Value_by_Element_Unit
sheet_plots$Value_by_Element_Unit <- {
  d <- value_by_element_unit %>%
    dplyr::mutate(grp = paste(element, "-", unit))
  ggplot2::ggplot(d, ggplot2::aes(stats::reorder(grp, n), n, fill = element)) +
    ggplot2::geom_col(width = .7) +
    ggplot2::geom_text(ggplot2::aes(label = paste0(scales::comma(n), "\nmedian ",
                                                   scales::comma(round(median)))),
                       hjust = -0.08, size = 3, lineheight = .95) +
    ggplot2::coord_flip() +
    ggplot2::scale_y_continuous(labels = scales::comma,
                                expand = ggplot2::expansion(mult = c(0, .28))) +
    ggplot2::scale_fill_manual(values = c("Import Quantity" = unname(pal["main"]),
                                          "Export Quantity" = unname(pal["alt"])),
                               name = NULL) +
    ggplot2::labs(title = "Records and typical size by element and unit",
                  subtitle = "Head (An), thousand head (1000 An) and tonnes (t) restate the same shipments",
                  x = NULL, y = "Records") +
    theme_desc()
}

## 10. Value_by_Year - distribution of the reported value through time
sheet_plots$Value_by_Year <- {
  d <- value_by_year %>%
    dplyr::select(year, q1, median, q3, mean) %>%
    tidyr::pivot_longer(c(median, mean), names_to = "stat", values_to = "value") %>%
    dplyr::mutate(stat = dplyr::recode(stat,
      median = "Median (with inter-quartile band)", mean = "Mean"))
  ggplot2::ggplot(d, ggplot2::aes(year, value)) +
    ggplot2::geom_ribbon(data = dplyr::filter(d, stat != "Mean"),
                         ggplot2::aes(ymin = q1, ymax = q3),
                         fill = pal["main"], alpha = .20) +
    ggplot2::geom_line(ggplot2::aes(colour = stat), linewidth = .9) +
    ggplot2::geom_point(ggplot2::aes(colour = stat), size = 1.9) +
    ggplot2::facet_wrap(~ stat, scales = "free_y", ncol = 1) +
    ggplot2::scale_x_continuous(breaks = value_by_year$year) +
    ggplot2::scale_y_continuous(labels = scales::comma) +
    ggplot2::scale_colour_manual(values = stats::setNames(
      c(unname(pal["alt"]), unname(pal["main"])),
      c("Mean", "Median (with inter-quartile band)")), guide = "none") +
    ggplot2::labs(title = "Reported quantity per record, by year",
                  subtitle = "All units pooled - the mean is pulled up by a few very large poultry consignments",
                  x = NULL, y = "Value per record") +
    theme_desc()
}

## 11. Head_by_Species
sheet_plots$Head_by_Species <- {
  d <- head_by_species %>%
    dplyr::mutate(item = stats::reorder(item, sum))
  ggplot2::ggplot(d, ggplot2::aes(sum, item)) +
    ggplot2::geom_col(fill = pal["main"], width = .72) +
    ggplot2::geom_text(ggplot2::aes(label = paste0(scales::label_number(scale_cut =
                         scales::cut_short_scale())(sum), "  (", scales::comma(n), " mov.)")),
                       hjust = -0.06, size = 3) +
    ggplot2::scale_x_continuous(labels = scales::label_number(scale_cut =
                                  scales::cut_short_scale()),
                                expand = ggplot2::expansion(mult = c(0, .3))) +
    ggplot2::labs(title = "Animals moved by species, 2005-2021",
                  subtitle = "Total head over all movements; movement count in brackets",
                  x = "Head", y = NULL) +
    theme_desc()
}

## 12. Head_by_Year
sheet_plots$Head_by_Year <- {
  ggplot2::ggplot(head_by_year, ggplot2::aes(year, sum)) +
    ggplot2::geom_col(fill = pal["main"], width = .68) +
    ggplot2::geom_vline(xintercept = 2019, linetype = "22", colour = pal["warn"]) +
    ggplot2::annotate("text", x = 2019, y = max(head_by_year$sum) * .96,
                      label = "2019 not in the data", colour = pal["warn"],
                      angle = 90, vjust = -0.5, size = 3) +
    ggplot2::scale_x_continuous(breaks = head_by_year$year) +
    ggplot2::scale_y_continuous(labels = scales::label_number(scale_cut =
                                  scales::cut_short_scale())) +
    ggplot2::labs(title = "Total animals moved per year",
                  subtitle = "Sum of head over all species and all routes",
                  x = NULL, y = "Head") +
    theme_desc()
}

## 13. Year_Summary
sheet_plots$Year_Summary <- {
  d <- year_summary %>%
    dplyr::select(year, n_movements, n_origins, n_destinations, n_species) %>%
    tidyr::pivot_longer(-year, names_to = "metric", values_to = "value") %>%
    dplyr::mutate(metric = dplyr::recode(metric,
      n_movements = "Movements recorded", n_origins = "Exporting countries",
      n_destinations = "Importing countries", n_species = "Species traded"))
  ggplot2::ggplot(d, ggplot2::aes(year, value, colour = metric)) +
    ggplot2::geom_vline(xintercept = 2019, linetype = "22", colour = "grey65") +
    ggplot2::geom_line(linewidth = .9) + ggplot2::geom_point(size = 1.8) +
    ggplot2::facet_wrap(~ metric, scales = "free_y", ncol = 2) +
    ggplot2::scale_x_continuous(breaks = seq(2005, 2021, 4)) +
    ggplot2::scale_y_continuous(labels = scales::comma) +
    ggplot2::scale_colour_brewer(palette = "Dark2", guide = "none") +
    ggplot2::labs(title = "Coverage of the network through time",
                  subtitle = "Dashed line marks 2019, which is absent from the source files - the series joins 2018 to 2020",
                  x = NULL, y = NULL) +
    theme_desc()
}

## 14. Country_Summary
sheet_plots$Country_Summary <- {
  d <- country_summary %>%
    dplyr::mutate(total = out_movements + in_movements) %>%
    dplyr::slice_max(total, n = 20) %>%
    dplyr::mutate(country = lab_of(country),
                  country = stats::reorder(country, total)) %>%
    dplyr::select(country, `Movements out` = out_movements,
                  `Movements in` = in_movements) %>%
    tidyr::pivot_longer(-country, names_to = "direction", values_to = "n")
  ggplot2::ggplot(d, ggplot2::aes(n, country, fill = direction)) +
    ggplot2::geom_col(width = .74) +
    ggplot2::scale_fill_manual(values = c("Movements out" = unname(pal["alt"]),
                                          "Movements in"  = unname(pal["main"])), name = NULL) +
    ggplot2::scale_x_continuous(labels = scales::comma) +
    ggplot2::labs(title = "Twenty busiest countries",
                  subtitle = "Movements recorded as origin and as destination, 2005-2021",
                  x = "Movements", y = NULL) +
    theme_desc()
}

## 15. Movement_Reporting
sheet_plots$Movement_Reporting <- {
  d <- janitor_tbl %>%
    dplyr::mutate(pct = n_movements / sum(n_movements),
                  reported_by = stats::reorder(reported_by, n_movements))
  ggplot2::ggplot(d, ggplot2::aes(n_movements, reported_by, fill = reported_by)) +
    ggplot2::geom_col(width = .62) +
    ggplot2::geom_text(ggplot2::aes(label = paste0(scales::comma(n_movements), "  (",
                                                   scales::percent(pct, accuracy = .1), ")")),
                       hjust = -0.06, size = 3.4) +
    ggplot2::scale_x_continuous(labels = scales::comma,
                                expand = ggplot2::expansion(mult = c(0, .3))) +
    ggplot2::scale_fill_manual(values = c("both sides" = unname(pal["ok"]),
                                          "exporter only" = unname(pal["alt"]),
                                          "importer only" = unname(pal["main"])),
                               guide = "none") +
    ggplot2::labs(title = "Who reported each movement",
                  subtitle = "Nearly half of all movements were reported by both trading partners and had to be collapsed",
                  x = "Movements", y = NULL) +
    theme_desc()
}

## 16. Network_Summary_All
sheet_plots$Network_Summary_All <- {
  d <- stats_all %>%
    dplyr::select(-network) %>%
    tidyr::pivot_longer(dplyr::everything(), names_to = "metric", values_to = "value") %>%
    dplyr::mutate(
      panel = ifelse(metric %in% c("density", "reciprocity", "transitivity"),
                     "Proportions (0-1)", "Counts"),
      metric = dplyr::recode(metric,
        n_nodes = "Countries", n_edges = "Routes", density = "Density",
        reciprocity = "Reciprocity", transitivity = "Transitivity",
        mean_in_degree = "Mean in-degree", mean_out_degree = "Mean out-degree",
        max_in_degree = "Max in-degree", max_out_degree = "Max out-degree",
        diameter = "Diameter", n_components_weak = "Weak components"))
  ggplot2::ggplot(d, ggplot2::aes(value, stats::reorder(metric, value), colour = panel)) +
    ggplot2::geom_segment(ggplot2::aes(x = 0.1, xend = value, yend = metric),
                          linewidth = .9) +
    ggplot2::geom_point(size = 3.6) +
    ggplot2::geom_text(ggplot2::aes(label = ifelse(value < 1, round(value, 3),
                                                   scales::comma(round(value, 1)))),
                       hjust = -0.35, size = 3.2, colour = "grey20") +
    ggplot2::scale_x_log10(labels = scales::comma,
                           expand = ggplot2::expansion(mult = c(0.02, .22))) +
    ggplot2::scale_colour_manual(values = c("Counts" = unname(pal["main"]),
                                            "Proportions (0-1)" = unname(pal["alt"])),
                                 name = NULL) +
    ggplot2::labs(title = "Structure of the combined network",
                  subtitle = "All species, all years - log scale, so proportions and counts fit on one axis",
                  x = "Value (log scale)", y = NULL) +
    theme_desc()
}

## 17. Network_Summary_Spp
sheet_plots$Network_Summary_Spp <- {
  ggplot2::ggplot(species_stats,
                  ggplot2::aes(n_nodes, n_edges, size = density, colour = reciprocity)) +
    ggplot2::geom_point(alpha = .85) +
    ggrepel::geom_text_repel(ggplot2::aes(label = network), size = 3,
                             colour = "grey20", seed = 1, max.overlaps = 30) +
    ggplot2::scale_size_continuous(range = c(2, 10), name = "Density") +
    ggplot2::scale_colour_gradient(low = "#a8dadc", high = "#1d3557", name = "Reciprocity") +
    ggplot2::scale_y_continuous(labels = scales::comma) +
    ggplot2::labs(title = "Size and shape of each species network",
                  subtitle = "Countries involved vs routes used; point size = density",
                  x = "Countries in the network", y = "Directed routes") +
    theme_desc()
}

## 18. Centrality_All
sheet_plots$Centrality_All <- {
  d <- cent_all %>% dplyr::mutate(lbl = lab_of(country))
  top <- d %>% dplyr::slice_max(total_degree, n = 14, with_ties = FALSE)
  ggplot2::ggplot(d, ggplot2::aes(out_degree, in_degree)) +
    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "22", colour = "grey70") +
    ggplot2::geom_point(ggplot2::aes(colour = total_degree, size = total_degree), alpha = .85) +
    ggrepel::geom_text_repel(data = top, ggplot2::aes(label = lbl), size = 3,
                             colour = "grey15", seed = 1, max.overlaps = 30) +
    ggplot2::scale_colour_gradient(low = "#ffd166", high = "#7f1d3f", name = "Total degree") +
    ggplot2::scale_size_continuous(range = c(1.2, 7), guide = "none") +
    ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = .09)) +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = .09)) +
    ggplot2::labs(title = "Exporting reach against importing reach",
                  subtitle = "Each point is a country; above the dashed line = imports from more partners than it exports to",
                  x = "Out-degree (countries supplied)", y = "In-degree (countries supplied by)") +
    theme_desc()
}

## 19. Centrality_by_Species
sheet_plots$Centrality_by_Species <- {
  top_c <- cent_all %>% dplyr::slice_max(total_degree, n = 20, with_ties = FALSE) %>%
    dplyr::pull(country)
  d <- species_cent %>%
    dplyr::filter(country %in% top_c) %>%
    dplyr::mutate(lbl = lab_of(country))
  ord <- d %>% dplyr::group_by(lbl) %>% dplyr::summarise(s = sum(total_degree)) %>%
    dplyr::arrange(s) %>% dplyr::pull(lbl)
  sp_ord <- d %>% dplyr::group_by(species) %>% dplyr::summarise(s = sum(total_degree)) %>%
    dplyr::arrange(dplyr::desc(s)) %>% dplyr::pull(species)
  d <- d %>% dplyr::mutate(lbl = factor(lbl, levels = ord),
                           species = factor(species, levels = sp_ord))
  ggplot2::ggplot(d, ggplot2::aes(species, lbl, fill = total_degree)) +
    ggplot2::geom_tile(colour = "white", linewidth = .4) +
    ggplot2::scale_fill_gradient(low = "#f1faee", high = "#1d3557", name = "Total degree") +
    ggplot2::labs(title = "Total degree of the twenty most connected countries, by species",
                  subtitle = "Darker = trades that species with more partners",
                  x = NULL, y = NULL) +
    theme_desc() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                   panel.grid = ggplot2::element_blank())
}

## 20. Top_Edges_All
sheet_plots$Top_Edges_All <- {
  d <- edges_all %>%
    dplyr::slice_max(weight, n = 20, with_ties = FALSE) %>%
    dplyr::mutate(route = paste(lab_of(from), "->", lab_of(to)),
                  route = stats::reorder(route, weight))
  ggplot2::ggplot(d, ggplot2::aes(weight, route, fill = n_species)) +
    ggplot2::geom_col(width = .74) +
    ggplot2::geom_text(ggplot2::aes(label = weight), hjust = -0.25, size = 3) +
    ggplot2::scale_fill_gradient(low = "#a8dadc", high = "#1d3557", name = "Species on route") +
    ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0, .12))) +
    ggplot2::labs(title = "Twenty most frequently used routes",
                  subtitle = "Weight = number of distinct species-year movements on that route",
                  x = "Movement frequency", y = NULL) +
    theme_desc()
}

stopifnot(setequal(names(sheet_plots), names(wb_sheets)))

# chart size per sheet (inches)
chart_size <- list(
  Categorical_Freq      = c(11, 7.5),
  Year_Summary          = c(9, 6),
  Centrality_by_Species = c(9.5, 7),
  Network_Summary_All   = c(10, 5),
  File_Inventory        = c(9.5, 5.5),
  Variable_Overview     = c(9.5, 5.5),
  Country_Summary       = c(9, 6)
)
size_of <- function(nm) if (!is.null(chart_size[[nm]])) chart_size[[nm]] else c(9, 5)

chart_files <- vapply(names(sheet_plots), function(nm) {
  wh <- size_of(nm)
  f  <- file.path(chart_dir, paste0(nm, ".png"))
  ggplot2::ggsave(f, sheet_plots[[nm]], width = wh[1], height = wh[2], dpi = 150, bg = "white")
  f
}, character(1))
```

## Writing the workbook

Each sheet holds its table on the left and its figure to the right of the data.

```{r export-xlsx-write}
xlsx_path <- file.path(tab_dir, "Descriptive_Statistics.xlsx")

wb  <- openxlsx::createWorkbook()
hdr <- openxlsx::createStyle(textDecoration = "bold", fgFill = "#1f6f8b",
                             fontColour = "white", halign = "left", border = "bottom")

for (nm in names(wb_sheets)) {
  df <- wb_sheets[[nm]]
  openxlsx::addWorksheet(wb, nm)
  openxlsx::writeData(wb, nm, df, headerStyle = hdr)
  openxlsx::freezePane(wb, nm, firstRow = TRUE)
  openxlsx::setColWidths(wb, nm, cols = seq_len(ncol(df)),
                         widths = pmin(60, pmax(10, nchar(names(df)) + 4)))
  wh <- size_of(nm)
  openxlsx::insertImage(wb, nm, chart_files[[nm]],
                        width = wh[1], height = wh[2],
                        startCol = ncol(df) + 2, startRow = 2)
}

openxlsx::saveWorkbook(wb, xlsx_path, overwrite = TRUE)

cat("Workbook written to", xlsx_path, "\n",
    length(wb_sheets), "sheets, each with its own chart:\n  ",
    paste(names(wb_sheets), collapse = ", "), "\n")
```

```{r session}
sessionInfo()
```
